Imputation of Incomplete Multilevel Data with R

### 2023-12-08 docs/manuscript.qmd
Add introduction text

---
title: "Imputation of Incomplete Multilevel Data with R"
title-formatted: "Imputation of Incomplete Multilevel Data with [R]{.proglang}"
format:
    jss-pdf:
        keep-tex: true
        fig-pos: 't'
        fig-width: 7
        fig-height: 3.5
    jss-html: default
author:
  - name: Hanne I. Oberman
    affiliations:
      - name: Utrecht University
        department: Methodology and Statistics
        address: Padualaan 14
        city: Utrecht
        country: The Netherlands
        postal-code: 3584 CH
      - Journal of Statistical Software
    orcid: 0000-0003-3276-2141
    email: h.i.oberman@uu.nl
    url: https://www.hanneoberman.github.io
  - name: Johanna Muñoz
    affiliations:
      - name: University Medical Center Utrecht
        department: Julius Centre for Health Sciences and Primary Care
        address: Universiteitsweg 100
        city: Utrecht
        country: The Netherlands
        postal-code: 3584 CG
      - Journal of Statistical Software
    orcid: 0000-0002-2384-5415
  - name: Valentijn M.T. de Jong
    affiliations:
      - name: University Medical Center Utrecht
        department: Julius Centre for Health Sciences and Primary Care
        city: Utrecht
        country: The Netherlands
      - name: European Medicines Agency
        department: Data Analytics and Methods Task Force
        city: Amsterdam
        country: The Netherlands
      - Journal of Statistical Software
    orcid: 0000-0001-9921-3468
  - name: Gerko Vink
    affiliations:
      - name: University Medical Center Utrecht
        department: Julius Centre for Health Sciences and Primary Care
        address: Universiteitsweg 100
        city: Utrecht
        country: The Netherlands
        postal-code: 3584 CG
      - Journal of Statistical Software
    orcid: 0000-0001-9767-1924
  - name: Thomas P.A. Debray
    affiliations:
      - name: University Medical Center Utrecht
        department: Julius Centre for Health Sciences and Primary Care
        address: Universiteitsweg 100
        city: Utrecht
        country: The Netherlands
        postal-code: 3584 CG
      - Journal of Statistical Software
    orcid: 0000-0002-1790-2719    
abstract: |
  This tutorial illustrates the imputation of incomplete multilevel data with the [R]{.proglang} packackage [mice]{.pkg}. Our scope is only simple multilevel models, to show how imputation can yield less biased estimates from incomplete clustered data. More complex models can be accomodated, but are outside the scope of this paper. Incomplete multilevel data requires careful consideration of the missing data problem and analysis strategy. In this tutorial, we focus on a popular strategy for accommodating missingness in multilevel data: replacing the missing data with one or more plausible values, i.e., imputation.Imputation separates the missing data problem from the main analysis and the completed data can be analyzed as if it has been fully observed. This tutorial illustrates the imputation of incomplete multilevel data with the statistical programming language R. We aim to show how imputation can yield less biased estimates from incomplete clustered data. We provide practical guidelines and code snippets for different missing data situations, including non-ignorable missingness mechanisms. For brevity, we focus on multilevel imputation using chained equations with the R mice package and its adjacent packages.

keywords: [missing data, multilevel, clustering, mice, R]
keywords-formatted: [missing data, multilevel, clustering, "[mice]{.pkg}", "[R]{.proglang}"]

bibliography: bibliography.bib  
---

@@

@@

The popular [mice]{.pkg} package in [R]{.proglang} @R...

@@

@@

@@

@@

@@

::: callout
Box 2. The predictor matrix.
:::

>>>>

---
title: "Imputation of Incomplete Multilevel Data with R"
title-formatted: "Imputation of Incomplete Multilevel Data with [R]{.proglang}"
format:
    jss-pdf:
        keep-tex: true
        fig-pos: 't'
        fig-width: 7
        fig-height: 3.5
    jss-html: default
author:
  - name: Hanne I. Oberman
    affiliations:
      - name: Utrecht University
        department: Methodology and Statistics
        address: Padualaan 14
        city: Utrecht
        country: The Netherlands
        postal-code: 3584 CH
      - Journal of Statistical Software
    orcid: 0000-0003-3276-2141
    email: h.i.oberman@uu.nl
    url: https://www.hanneoberman.github.io
  - name: Johanna Muñoz
    affiliations:
      - name: University Medical Center Utrecht
        department: Julius Centre for Health Sciences and Primary Care
        address: Universiteitsweg 100
        city: Utrecht
        country: The Netherlands
        postal-code: 3584 CG
      - Journal of Statistical Software
    orcid: 0000-0002-2384-5415
  - name: Valentijn M.T. de Jong
    affiliations:
      - name: University Medical Center Utrecht
        department: Julius Centre for Health Sciences and Primary Care
        city: Utrecht
        country: The Netherlands
      - name: European Medicines Agency
        department: Data Analytics and Methods Task Force
        city: Amsterdam
        country: The Netherlands
      - Journal of Statistical Software
    orcid: 0000-0001-9921-3468
  - name: Gerko Vink
    affiliations:
      - name: University Medical Center Utrecht
        department: Julius Centre for Health Sciences and Primary Care
        address: Universiteitsweg 100
        city: Utrecht
        country: The Netherlands
        postal-code: 3584 CG
      - Journal of Statistical Software
    orcid: 0000-0001-9767-1924
  - name: Thomas P.A. Debray
    affiliations:
      - name: University Medical Center Utrecht
        department: Julius Centre for Health Sciences and Primary Care
        address: Universiteitsweg 100
        city: Utrecht
        country: The Netherlands
        postal-code: 3584 CG
      - Journal of Statistical Software
    orcid: 0000-0002-1790-2719    
abstract: |
  This tutorial illustrates the imputation of incomplete multilevel data with the [R]{.proglang} packackage [mice]{.pkg}. Our scope is only simple multilevel models, to show how imputation can yield less biased estimates from incomplete clustered data. More complex models can be accomodated, but are outside the scope of this paper. Incomplete multilevel data requires careful consideration of the missing data problem and analysis strategy. In this tutorial, we focus on a popular strategy for accommodating missingness in multilevel data: replacing the missing data with one or more plausible values, i.e., imputation.Imputation separates the missing data problem from the main analysis and the completed data can be analyzed as if it has been fully observed. This tutorial illustrates the imputation of incomplete multilevel data with the statistical programming language R. We aim to show how imputation can yield less biased estimates from incomplete clustered data. We provide practical guidelines and code snippets for different missing data situations, including non-ignorable missingness mechanisms. For brevity, we focus on multilevel imputation using chained equations with the R mice package and its adjacent packages.
keywords: [missing data, multilevel, clustering, mice, R]
keywords-formatted: [missing data, multilevel, clustering, "[mice]{.pkg}", "[R]{.proglang}"]
bibliography: bibliography.bib  
---

@@

```{r}
#| label: setup
#| echo: false
library(ggplot2)
source("../docs/utils.R")
```

@@

The popular [mice]{.pkg} package in [R]{.proglang} @R

@@

This papers serves as a tutorial for imputing incomplete multilevel data with [mice]{.pkg} in [R]{.proglang}. [mice]{.pkg} has become the de-facto standard for imputation by chained equations, which iteratively solves the missingness on a variable-by-variable basis. [mice]{.pkg} is known to yield valid inferences under many different missing data circumstances [@buur18].

We provide practical guidelines and code snippets for different missing data situations, including non-ignorable mechanisms. For reasons of brevity, we focus on multilevel imputation by chained equations with [mice]{.pkg} exclusively; other imputation methods and packages [see e.g. @audi18 and @grun18] are outside the scope of this tutorial. Assumed knowledge includes basic familiarity with the [lme4]{.pkg} notation for multilevel models (see Table \ref{tab:mod}).

We illustrate imputation of incomplete multilevel data using three case studies:

- `popmis` from the [mice]{.pkg} package [simulated data on perceived popularity, $n = 2,000$ pupils across $N = 100$ schools with data that are MAR, @mice];
- `impact` from the [metamisc]{.pkg} package [empirical data on traumatic brain injuries, $n = 11,022$ patients across $N = 15$ studies with data that are MAR, @metamisc];
- `obesity` from the [micemd]{.pkg} package [simulated data on obesity, $n = 2,111$ patients across $N = 5$ regions with data that are MNAR].

For each of these datasets, we discuss the nature of the missingness, choose one or more imputation models and evaluate the imputed data, but we will also highlight one specific aspect of the imputation workflow.

This tutorial is dedicated to readers who are unfamiliar with multiple imputation. More experienced readers can skip the introduction (case study 1) and directly head to practical applications of multilevel imputation under MAR conditions (case study 2) or under MNAR conditions (case study 3).

@@

Many datasets include individuals that are clustered together, for example in geographic regions, or even different studies. In the simplest case, individuals (e.g., students) are nested within a single cluster (e.g., school classes). More complex clustered structures may occur when there are multiple hierarchical levels (e.g., students in different schools or patients within hospitals within regions across countries), or when the clustering is non-nested (e.g., electronic health record data from diverse settings and populations within large databases). With clustered data we generally assume that individuals from the same cluster tend to be more similar than individuals from other clusters. In statistical terms, this implies that observations from the same cluster are not independent and may in fact be correlated. If this correlation is left unaddressed, estimates of *p* values, confidence intervals even model parameters are prone to bias  [@loca01]. Statistical methods for clustered data typically adopt hierarchical models that explicitly describe the grouping of observations. These models are also known as 'multilevel models', 'hierarchical models', 'mixed effect models', 'random effect models', and in the context of time-to-event data as 'frailty models'. Table \ref{tab:clus} provides an overview of some key concepts in multilevel modeling.

@@

```{R}
#| echo: false
concepts <-
  data.frame(
    Concept = c(
      "Sample units",
      "Cluster",
      "Hierarchical data",
      "Level-1",
      "Level-2",
      "Hierarchical model",
      "Fixed effect",
      "Random effect",
      "Mixed effect",
      "ICC",
      "Stratified intercept"
    ),
    Details = c(
      "Units of the population from which measurements are taken in a sample, e.g., students.",
      "Variable that specify the cluster or agruppation, e.g., Classroom",
      "Data are grouped into clusters at different levels, observations belonging to the same cluster are expected to share certain characteristics.",
      "Variable that varies within a cluster, eg. Test score",
      "Variable that does not vary within a cluster but between, e.g. teacher experience.",
      "Model accounting for dependant observations relying on certain parameters ( within cluster) which in turn depend on other parameters (between cluster)" ,
      "Effects that are constant across all sample units, e.g. something that researchers control for and can repeat, such as  a teaching strategy (tutoring after class)",
      "Effects that are a source of random variation in the data, and whose levels are not fully sampled. e.g. test score tendency during academic year between students due to no controlled factors such as  genetic,family history",
      "Includes fixed and random effects, e.g. the fixed effect would be the treatment effect of a drug and the random effect would be the ID of the hospital where the patient is treated. Multilevel models typically accommodate for variability by including a separate group mean for each cluster e.g random intercept on hospitals. In addition to random intercepts, multilevel models can also include random coefficients and heterogeneous residual error variances across clusters (see e.g. @gelm06, @hox17 and @jong21).",
      "The variability due to clustering is often measured by means of the intraclass coefficient (ICC). The ICC can be seen as the percentage
of variance that can be attributed to the cluster-level, where a high ICC would indicate that a lot of variability is due to the cluster structure.",
""
    )
  )
# concepts |>
#   kable(
#     booktabs = T,
#     escape = F,
#     caption = "Concepts in multilevel methods",
#     col.names = c("Concept", "Details"),
#     align = "l"
#   ) |>
#   column_spec(c(1), width = "3cm") |>
#   column_spec(c(2), width = "12cm")
```

@@

missing data mechanisms etc.

As with any other dataset, clustered datasets may be impacted by missingness in much the same way. Several strategies can be used to handle missing data, including complete case analysis and imputation. We focus on the latter approach and discuss statistical methods for replacing the missing data with one or more plausible values. Imputation separates the missing data problem from the analysis and the completed data can be analyzed as if it were completely observed. It is generally recommended to impute the missing values more than once to preserve uncertainty due to missingness and to allow for valid inferences (c.f. Rubin 1976).

With incomplete clustered datasets we can distinguish between two types of missing data: sporadic missingness and systematic missingness [@resc13]. Sporadic missingness arises when variables are missing for some but not all of the units in a cluster [@buur18; @jola18]. For example, it is possible that test results are missing for several students in one or more classes. When all observations are missing within one or more clusters, data are said to be systematically missing. Sporadic missingness is visualized in Figure XYZ.

```{R}
#| echo: false
plot_na()
```

Column $X_1$ in Figure 1 is completely observed, column $X_2$ is systematically missing in cluster 2, and column $X_3$ is sporadically missing. To analyze these incomplete data, we have to take the nature of the missingness and the cluster structure into account. For example, the sporadic missingness in $X_3$ could be easily amended if this would be a cluster-level variable (and thus constant within clusters). We could then just extrapolate the true (but missing) value of $X_3$ for unit 1 from unit 2, and the value for unit 4 from unit 3. If $X_3$ would instead be a unit-level variable (which may vary within clusters), we could not just recover the unobserved 'truth', but would need to use some kind of missing data method, or discard the incomplete units altogether (i.e., complete case analysis). Complete case analysis can however introduce bias in statistical inferences and lowers statistical power. Further, with the systematic missingness in $X_2$, it would be impossible to fit a multilevel model without accommodating the missingness in some way. Complete case analysis in that case would mean excluding the entire cluster from the analyses. The wrong choice of missing data handling method can thus be extremely harmful to the inferences.

Imputation of missing data requires consideration of the mechanism behind the missingness. Rubin proposed to distinguish between data that are missing completely at random (MCAR), data that are missing at random (MAR) and data that are missing not at random (MNAR; see Table \ref{tab:miss}). For each of these three missingness generating mechanisms, different imputation strategies are warranted (@yuce08 and @hox15). We here consider the general case that data are MAR, and expand on certain MNAR situations.

### imputation with mice

@@

::: callout
Box 3. The predictor matrix. The entries corresponding to the level-1 predictors are coded with a 3, indicating that both the original values as well as the cluster means of the predictor are included into the imputation model. The entry of 4 in the predictor matrix adds three variables to the imputation model for the imputation model predictor: the value of the predictor, the cluster means of the predictor and the random slopes of the predictor.
:::

### 2023-12-08 docs/manuscript.tex
Add introduction text

@@

The popular \pkg{mice} package in \proglang{R} \citet{R}\ldots{}

@@

@@

\begin{tcolorbox}[enhanced jigsaw, left=2mm, leftrule=.75mm, bottomrule=.15mm, toprule=.15mm, breakable, colback=white, opacityback=0, arc=.35mm, rightrule=.15mm]

@@

@@

\begin{tcolorbox}[enhanced jigsaw, left=2mm, leftrule=.75mm, bottomrule=.15mm, toprule=.15mm, breakable, colback=white, opacityback=0, arc=.35mm, rightrule=.15mm]

@@

\begin{tcolorbox}[enhanced jigsaw, left=2mm, leftrule=.75mm, bottomrule=.15mm, toprule=.15mm, breakable, colback=white, opacityback=0, arc=.35mm, rightrule=.15mm]

Box 2. The predictor matrix.

@@

{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-4-1.pdf}

@@

@@

{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-12-1.pdf}

@@

{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-14-1.pdf}

@@

@@

{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-16-1.pdf}

@@

>>>>

\begin{verbatim}
Warning: package 'ggplot2' was built under R version 4.3.2
\end{verbatim}

@@

The popular \pkg{mice} package in \proglang{R} \citet{R}

@@

This papers serves as a tutorial for imputing incomplete multilevel data
with \pkg{mice} in \proglang{R}. \pkg{mice} has become the de-facto
standard for imputation by chained equations, which iteratively solves
the missingness on a variable-by-variable basis. \pkg{mice} is known to
yield valid inferences under many different missing data circumstances
\citep{buur18}.

We provide practical guidelines and code snippets for different missing
data situations, including non-ignorable mechanisms. For reasons of
brevity, we focus on multilevel imputation by chained equations with
\pkg{mice} exclusively; other imputation methods and packages \citep[see
e.g.][ and \citet{grun18}]{audi18} are outside the scope of this
tutorial. Assumed knowledge includes basic familiarity with the
\pkg{lme4} notation for multilevel models (see Table \ref{tab:mod}).

We illustrate imputation of incomplete multilevel data using three case
studies:

\begin{itemize}
\tightlist
\item
  \texttt{popmis} from the \pkg{mice} package \citep[simulated data on
  perceived popularity, \(n = 2,000\) pupils across \(N = 100\) schools
  with data that are MAR,][]{mice};
\item
  \texttt{impact} from the \pkg{metamisc} package \citep[empirical data
  on traumatic brain injuries, \(n = 11,022\) patients across \(N = 15\)
  studies with data that are MAR,][]{metamisc};
\item
  \texttt{obesity} from the \pkg{micemd} package {[}simulated data on
  obesity, \(n = 2,111\) patients across \(N = 5\) regions with data
  that are MNAR{]}.
\end{itemize}

For each of these datasets, we discuss the nature of the missingness,
choose one or more imputation models and evaluate the imputed data, but
we will also highlight one specific aspect of the imputation workflow.

This tutorial is dedicated to readers who are unfamiliar with multiple
imputation. More experienced readers can skip the introduction (case
study 1) and directly head to practical applications of multilevel
imputation under MAR conditions (case study 2) or under MNAR conditions
(case study 3).

@@

Many datasets include individuals that are clustered together, for
example in geographic regions, or even different studies. In the
simplest case, individuals (e.g., students) are nested within a single
cluster (e.g., school classes). More complex clustered structures may
occur when there are multiple hierarchical levels (e.g., students in
different schools or patients within hospitals within regions across
countries), or when the clustering is non-nested (e.g., electronic
health record data from diverse settings and populations within large
databases). With clustered data we generally assume that individuals
from the same cluster tend to be more similar than individuals from
other clusters. In statistical terms, this implies that observations
from the same cluster are not independent and may in fact be correlated.
If this correlation is left unaddressed, estimates of \emph{p} values,
confidence intervals even model parameters are prone to bias
\citep{loca01}. Statistical methods for clustered data typically adopt
hierarchical models that explicitly describe the grouping of
observations. These models are also known as `multilevel models',
`hierarchical models', `mixed effect models', `random effect models',
and in the context of time-to-event data as `frailty models'. Table
\ref{tab:clus} provides an overview of some key concepts in multilevel
modeling.

\begin{tcolorbox}[enhanced jigsaw, arc=.35mm, bottomrule=.15mm, toprule=.15mm, breakable, leftrule=.75mm, left=2mm, colback=white, opacityback=0, rightrule=.15mm]

@@

missing data mechanisms etc.

As with any other dataset, clustered datasets may be impacted by
missingness in much the same way. Several strategies can be used to
handle missing data, including complete case analysis and imputation. We
focus on the latter approach and discuss statistical methods for
replacing the missing data with one or more plausible values. Imputation
separates the missing data problem from the analysis and the completed
data can be analyzed as if it were completely observed. It is generally
recommended to impute the missing values more than once to preserve
uncertainty due to missingness and to allow for valid inferences (c.f.
Rubin 1976).

With incomplete clustered datasets we can distinguish between two types
of missing data: sporadic missingness and systematic missingness
\citep{resc13}. Sporadic missingness arises when variables are missing
for some but not all of the units in a cluster \citep{buur18, jola18}.
For example, it is possible that test results are missing for several
students in one or more classes. When all observations are missing
within one or more clusters, data are said to be systematically missing.
Sporadic missingness is visualized in Figure XYZ.

\begin{verbatim}
plot_na()
\end{verbatim}

{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-3-1.pdf}

Column \(X_1\) in Figure 1 is completely observed, column \(X_2\) is
systematically missing in cluster 2, and column \(X_3\) is sporadically
missing. To analyze these incomplete data, we have to take the nature of
the missingness and the cluster structure into account. For example, the
sporadic missingness in \(X_3\) could be easily amended if this would be
a cluster-level variable (and thus constant within clusters). We could
then just extrapolate the true (but missing) value of \(X_3\) for unit 1
from unit 2, and the value for unit 4 from unit 3. If \(X_3\) would
instead be a unit-level variable (which may vary within clusters), we
could not just recover the unobserved `truth', but would need to use
some kind of missing data method, or discard the incomplete units
altogether (i.e., complete case analysis). Complete case analysis can
however introduce bias in statistical inferences and lowers statistical
power. Further, with the systematic missingness in \(X_2\), it would be
impossible to fit a multilevel model without accommodating the
missingness in some way. Complete case analysis in that case would mean
excluding the entire cluster from the analyses. The wrong choice of
missing data handling method can thus be extremely harmful to the
inferences.

Imputation of missing data requires consideration of the mechanism
behind the missingness. Rubin proposed to distinguish between data that
are missing completely at random (MCAR), data that are missing at random
(MAR) and data that are missing not at random (MNAR; see Table
\ref{tab:miss}). For each of these three missingness generating
mechanisms, different imputation strategies are warranted
(\citet{yuce08} and \citet{hox15}). We here consider the general case
that data are MAR, and expand on certain MNAR situations.

\hypertarget{imputation-with-mice}{%
\subsection{imputation with mice}\label{imputation-with-mice}}

@@

\begin{tcolorbox}[enhanced jigsaw, arc=.35mm, bottomrule=.15mm, toprule=.15mm, breakable, leftrule=.75mm, left=2mm, colback=white, opacityback=0, rightrule=.15mm]

@@

\begin{tcolorbox}[enhanced jigsaw, arc=.35mm, bottomrule=.15mm, toprule=.15mm, breakable, leftrule=.75mm, left=2mm, colback=white, opacityback=0, rightrule=.15mm]

Box 3. The predictor matrix. The entries corresponding to the level-1
predictors are coded with a 3, indicating that both the original values
as well as the cluster means of the predictor are included into the
imputation model. The entry of 4 in the predictor matrix adds three
variables to the imputation model for the imputation model predictor:
the value of the predictor, the cluster means of the predictor and the
random slopes of the predictor.

@@

@@

{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-10-1.pdf}

@@

@@

@@

{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-18-1.pdf}

@@

{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-19-1.pdf}

@@

{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-20-1.pdf}
